---
title: "HELIX meth-expr Cell adjusted results"
author: 
- Carlos Ruiz 

date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = "../")

```

# Introduction

In this report, we discuss the results of the association of methylation vs gene expression in the subcohort data of HELIX project. The results of this report were obtained by running a model adjusted for cell counts and only including autosome chromosomes.

```{r}
library(ggplot2)
library(dplyr)
load("results/MethComBatExpResidualsCellAdj/allres_simP_cpgs.Rdata")
load("data/cpganno.RData")

## Define useful variables
sigdf <- df[df$sigPair, ]
allCpGs <- unique(df$CpG)
sigCpGs <- unique(sigdf$CpG)
sigTCs <- unique(sigdf$TC)
```


# General overview

We evaluated the association between `r length(unique(df$CpG))` CpGs and `r length(unique(df$TC))` TCs. Of those, `r length(sigCpGs)` CpGs were significantly associated with at least one gene (`r round(length(sigCpGs)/length(unique(df$CpG))*100, 2)`%) and `r length(sigTCs)` were significantly associated with at least one CpG (`r round(length(sigTCs)/length(unique(df$TC))*100, 2)`).

In total, we tested `r nrow(df)` CpG-TC pairs. In `r nrow(sigdf)` of those pairs, changes in methylation were associated with changes in gene expression (`r round(nrow(sigdf)/nrow(df)*100, 2)`%). 

In general, each CpGs was associated with few TCs while each TCs was associated with several genes: 

```{r}
a <- data.frame(table(sigdf$CpG))
a$Feature <- "CpG"

b <- data.frame(table(sigdf$TC))
b$Feature <- "TC"
t <- rbind(a, b)
ggplot(t, aes(x = Freq)) + geom_histogram(binwidth = 1) + 
  ylab("Number of Features") + theme_bw() + facet_grid(~Feature) + 
  scale_x_continuous(limits = c(0, 30), name = "Feaures associated")


CpGs1 <- a[a$Freq == 1, "Var1"]
CpGs1m <- a[a$Freq != 1, "Var1"]

sigdf1 <- as_tibble(filter(sigdf, CpG %in% CpGs1))
sigdf1m <- as_tibble(filter(sigdf, CpG %in% CpGs1m))

sigdf1m_signSum <- sigdf1m %>% 
  group_by(CpG) %>% 
  summarize(propPos = mean(FC > 0), propNeg = mean(FC < 0))

```

`r sum(a$Freq == 1)` were only associated with one gene (`r round(mean(a$Freq == 1)*100, 2)`%), and we called them mono CpGs. On the other hand, we defined CpGs associated with more than 1 TC as pleiotropic CpGs. 

In mono CpGs, `r sum(sigdf1$FC > 0)` CpGs were positively associated with gene expression (`r round(mean(sigdf1$FC > 0)*100, 2)`%) and `r sum(sigdf1$FC < 0)` CpGs were negatively associated with gene expression (`r round(mean(sigdf1$FC < 0)*100, 2)`%). In multi CpGs, for `r sum(sigdf1m_signSum$propPos == 1)` CpGs the association was positive for all CpGs (`r round(mean(sigdf1m_signSum$propPos == 1)*100, 2)`%), for `r sum(sigdf1m_signSum$propNeg == 1)` CpGs the association was negative for all CpGs (`r round(mean(sigdf1m_signSum$propNeg == 1)*100, 2)`%) and `r sum(sigdf1m_signSum$propPos != 1 & sigdf1m_signSum$propNeg != 1)` CpGs were positively associated with some TCs and negatively associated with others (`r round(mean(sigdf1m_signSum$propPos != 1 & sigdf1m_signSum$propNeg != 1)*100, 2)`%). 

# CpG-TC distance




# CpGs Enrichment

In the following sections, we tested if the CpGs associated with gene expression had different features.

## Genic/Intergenic

We first tested whether CpGs associated with gene expression were located in genic or intergenic regions.

### Associated vs not-associated

```{r}
rownames(cpganno) <- cpganno$Name
allanno <- cpganno[allCpGs, ]
allanno$assoc <- allanno$Name %in% sigCpGs

allanno$GeneRel <- ifelse(allanno$UCSC_RefGene_Name == "", "Intergenic", "Genic")

t <- table(factor(allanno$assoc, levels = c("TRUE", "FALSE")), allanno$GeneRel)
t
chisq.test(t)

## OR
or <- round(t[1, 1]*t[2, 2]/(t[1, 2]*t[2, 1]), 2)
or
```

CpGs associated with gene expression tend to be more present in genic regions.


### Mono vs non-associated

```{r}
anno1 <- allanno[!allanno$Name %in% CpGs1m, ]
anno1$assoc <- anno1$Name %in% sigCpGs

t <- table(factor(anno1$assoc, levels = c("TRUE", "FALSE")), anno1$GeneRel)
t
chisq.test(t)

## OR
or <- round(t[1, 1]*t[2, 2]/(t[1, 2]*t[2, 1]), 2)
or
```

When we only consider mono CpGs, they are more focused on genic regions.


### Multi vs non-associated

```{r}
anno1m <- allanno[!allanno$Name %in% CpGs1, ]
anno1m$assoc <- anno1m$Name %in% sigCpGs

t <- table(factor(anno1m$assoc, levels = c("TRUE", "FALSE")), anno1m$GeneRel)
t
chisq.test(t)

## OR
or <- round(t[1, 1]*t[2, 2]/(t[1, 2]*t[2, 1]), 2)
or

```

When we only consider multi CpGs, they are mainly placed in intergenic regions.
