---
title: "Comparison methExprs adjustment P-value"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
      toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.wide = TRUE)
knitr::opts_knit$set(root.dir = "../")
```


```{r load}
library(dplyr)
library(UpSetR)

load("results/preprocessFiles/gexpAnnotation.Rdata")
expAnnot$TC <- expAnnot$transcript_cluster_id

```

# Cell counts adjusted

```{r}
load("results/MethComBatExpResidualsCellAdj/allres_simP_cpgs.Rdata")
modC <- df %>%
  as_tibble() %>%
  mutate(adj.p.value = p.adjust(p.value, "BH"),
         sigPair_BH = adj.p.value < 0.05,
         adj.p.value.BF = p.adjust(p.value, "bonferroni"),
         sigPair_BF = adj.p.value.BF < 0.05,
         pair = paste(CpG, TC))
featsC <- featStatsDF

```


## Pairs found

```{r}
permsP <- modC[modC$sigPair, ]$pair
BHP <- modC[modC$sigPair_BH, ]$pair
BFP <- modC[modC$sigPair_BF, ]$pair

upset(fromList(list(Permutation = permsP, BH = BHP, Bonferroni = BFP)), 
      order.by = "freq")
```

We found `r length(permsP)` pairs with the permutations, `r length(BHP)` pairs with BH and `r length(BFP)` with Bonferroni. 

## CpGs found

```{r}
permsCp <- unique(modC[modC$sigPair, ]$CpG)
BHCp <- unique(modC[modC$sigPair_BH, ]$CpG)
BFCp <- unique(modC[modC$sigPair_BF, ]$CpG)

upset(fromList(list(Permutation = permsCp, BH = BHCp, Bonferroni = BFCp)), 
      order.by = "freq")
```

We found `r length(permsCp)` CpGs with the permutations, `r length(BHCp)` CpGs with BH and `r length(BFCp)` with Bonferroni.  


```{r}
permsSum <- modC %>%
  group_by(CpG) %>%
  summarise(Type = ifelse(sum(sigPair) == 0, "Non-significant",
                          ifelse(sum(sigPair) == 1, "Mono", "Multi")),
            Direction = ifelse(sum(sigPair) == 0, "Non-significant",
                               ifelse(all(FC[sigPair] > 0), "Positive", 
                                      ifelse(all(FC[sigPair] < 0), "Negative", "Both"))))
BHSum <- modC %>%
  group_by(CpG) %>%
  summarise(Type = ifelse(sum(sigPair_BH) == 0, "Non-significant",
                          ifelse(sum(sigPair_BH) == 1, "Mono", "Multi")),
            Direction = ifelse(sum(sigPair_BH) == 0, "Non-significant",
                               ifelse(all(FC[sigPair_BH] > 0), "Positive", 
                                      ifelse(all(FC[sigPair_BH] < 0), "Negative", "Both"))))
BFSum <- modC %>%
  group_by(CpG) %>%
  summarise(Type = ifelse(sum(sigPair_BF) == 0, "Non-significant",
                          ifelse(sum(sigPair_BF) == 1, "Mono", "Multi")),
            Direction = ifelse(sum(sigPair_BF) == 0, "Non-significant",
                               ifelse(all(FC[sigPair_BF] > 0), "Positive", 
                                      ifelse(all(FC[sigPair_BF] < 0), "Negative", "Both"))))


```

### Single CpGs

```{r}
permsSing <- permsSum[permsSum$Type == "Mono", ]$CpG
BHCSing <- BHSum[BHSum$Type == "Mono", ]$CpG
BFCSing <- BFSum[BFSum$Type == "Mono", ]$CpG

upset(fromList(list(Permutation = permsSing, BH = BHCSing, Bonferroni = BFCSing)), 
      order.by = "freq")
```

We found `r length(permsSing)` single CpGs with the permutations, `r length(BHCSing)` single CpGs with BH and `r length(BFCSing)` with Bonferroni.  


### Multi CpGs

```{r}
permsMul <- permsSum[permsSum$Type == "Multi", ]$CpG
BHCMul <- BHSum[BHSum$Type == "Multi", ]$CpG
BFCMul <- BFSum[BFSum$Type == "Multi", ]$CpG

upset(fromList(list(Permutation = permsMul, BH = BHCMul, Bonferroni = BFCMul)), 
      order.by = "freq")
```

We found `r length(permsMul)` multi CpGs with the permutations, `r length(BHCMul)` multi CpGs with BH and `r length(BFCMul)` with Bonferroni.  

### Positive CpGs

```{r}
permsPos <- permsSum[permsSum$Direction == "Positive", ]$CpG
BHCPos <- BHSum[BHSum$Direction == "Positive", ]$CpG
BFCPos <- BFSum[BFSum$Direction == "Positive", ]$CpG

upset(fromList(list(Permutation = permsPos, BH = BHCPos, Bonferroni = BFCPos)), 
      order.by = "freq")
```

We found `r length(permsPos)` positive CpGs with the permutations, `r length(BHCPos)` positive CpGs with BH and `r length(BFCPos)` with Bonferroni.  

### Inverse CpGs

```{r}
permsNeg <- permsSum[permsSum$Direction == "Negative", ]$CpG
BHCNeg <- BHSum[BHSum$Direction == "Negative", ]$CpG
BFCNeg <- BFSum[BFSum$Direction == "Negative", ]$CpG

upset(fromList(list(Permutation = permsNeg, BH = BHCNeg, Bonferroni = BFCNeg)), 
      order.by = "freq")
```

We found `r length(permsNeg)` negative CpGs with the permutations, `r length(BHCNeg)` negative CpGs with BH and `r length(BFCNeg)` with Bonferroni.  



## TCs found

```{r}
permsTC <- unique(modC[modC$sigPair, ]$TC)
BHTC <- unique(modC[modC$sigPair_BH, ]$TC)
BFTC <- unique(modC[modC$sigPair_BF, ]$TC)

upset(fromList(list(Permutation = permsTC, BH = BHTC, Bonferroni = BFTC)), 
      order.by = "freq")
```

We found `r length(permsTC)` TCs with the permutations, `r length(BHTC)` TCs with BH and `r length(BFTC)` with Bonferroni. 


```{r}
modC <- modC %>%
  left_join(select(expAnnot, TC, Coding), by = "TC")

permsCodTC <- unique(modC[modC$sigPair & modC$Coding == "coding", ]$TC)
BHCodTC <- unique(modC[modC$sigPair_BH & modC$Coding == "coding", ]$TC)
BFCodTC <- unique(modC[modC$sigPair_BF & modC$Coding == "coding", ]$TC)

upset(fromList(list(Permutation = permsCodTC, BH = BHCodTC, Bonferroni = BFCodTC)), order.by = "freq")
```

We found `r length(permsCodTC)` coding TCs with the permutations, `r length(BHCodTC)` coding TCs with BH and `r length(BFCodTC)` with Bonferroni. 

# Variables Adjusted 

```{r}
load("results/MethComBatExpResidualsNoCellAdj/allres_simP_cpgs.Rdata")
modU <- df %>%
  as_tibble() %>%
  mutate(adj.p.value = p.adjust(p.value, "BH"),
         sigPair_BH = adj.p.value < 0.05,
         adj.p.value.BF = p.adjust(p.value, "bonferroni"),
         sigPair_BF = adj.p.value.BF < 0.05,
         pair = paste(CpG, TC))
featsU <- featStatsDF
```

## Pairs found

```{r}
permsP <- modU[modU$sigPair, ]$pair
BHP <- modU[modU$sigPair_BH, ]$pair
BFP <- modU[modU$sigPair_BF, ]$pair

upset(fromList(list(Permutation = permsP, BH = BHP, Bonferroni = BFP)), 
      order.by = "freq")
```

We found `r length(permsP)` pairs with the permutations, `r length(BHP)` pairs with BH and `r length(BFP)` with Bonferroni. 

## CpGs found

```{r}
permsCp <- unique(modU[modU$sigPair, ]$CpG)
BHCp <- unique(modU[modU$sigPair_BH, ]$CpG)
BFCp <- unique(modU[modU$sigPair_BF, ]$CpG)

upset(fromList(list(Permutation = permsCp, BH = BHCp, Bonferroni = BFCp)), 
      order.by = "freq")
```

We found `r length(permsCp)` CpGs with the permutations, `r length(BHCp)` CpGs with BH and `r length(BFCp)` with Bonferroni.  



```{r}
permsSum <- modU %>%
  group_by(CpG) %>%
  summarise(Type = ifelse(sum(sigPair) == 0, "Non-significant",
                          ifelse(sum(sigPair) == 1, "Mono", "Multi")),
            Direction = ifelse(sum(sigPair) == 0, "Non-significant",
                               ifelse(all(FC[sigPair] > 0), "Positive", 
                                      ifelse(all(FC[sigPair] < 0), "Negative", "Both"))))
BHSum <- modU %>%
  group_by(CpG) %>%
  summarise(Type = ifelse(sum(sigPair_BH) == 0, "Non-significant",
                          ifelse(sum(sigPair_BH) == 1, "Mono", "Multi")),
            Direction = ifelse(sum(sigPair_BH) == 0, "Non-significant",
                               ifelse(all(FC[sigPair_BH] > 0), "Positive", 
                                      ifelse(all(FC[sigPair_BH] < 0), "Negative", "Both"))))
BFSum <- modU %>%
  group_by(CpG) %>%
  summarise(Type = ifelse(sum(sigPair_BF) == 0, "Non-significant",
                          ifelse(sum(sigPair_BF) == 1, "Mono", "Multi")),
            Direction = ifelse(sum(sigPair_BF) == 0, "Non-significant",
                               ifelse(all(FC[sigPair_BF] > 0), "Positive", 
                                      ifelse(all(FC[sigPair_BF] < 0), "Negative", "Both"))))


```

### Single CpGs

```{r}
permsSing <- permsSum[permsSum$Type == "Mono", ]$CpG
BHCSing <- BHSum[BHSum$Type == "Mono", ]$CpG
BFCSing <- BFSum[BFSum$Type == "Mono", ]$CpG

upset(fromList(list(Permutation = permsSing, BH = BHCSing, Bonferroni = BFCSing)), 
      order.by = "freq")
```

We found `r length(permsSing)` single CpGs with the permutations, `r length(BHCSing)` single CpGs with BH and `r length(BFCSing)` with Bonferroni.  


### Multi CpGs

```{r}
permsMul <- permsSum[permsSum$Type == "Multi", ]$CpG
BHCMul <- BHSum[BHSum$Type == "Multi", ]$CpG
BFCMul <- BFSum[BFSum$Type == "Multi", ]$CpG

upset(fromList(list(Permutation = permsMul, BH = BHCMul, Bonferroni = BFCMul)),       order.by = "freq")
```

We found `r length(permsMul)` multi CpGs with the permutations, `r length(BHCMul)` multi CpGs with BH and `r length(BFCMul)` with Bonferroni.  

### Positive CpGs

```{r}
permsPos <- permsSum[permsSum$Direction == "Positive", ]$CpG
BHCPos <- BHSum[BHSum$Direction == "Positive", ]$CpG
BFCPos <- BFSum[BFSum$Direction == "Positive", ]$CpG

upset(fromList(list(Permutation = permsPos, BH = BHCPos, Bonferroni = BFCPos)),       order.by = "freq")
```

We found `r length(permsPos)` positive CpGs with the permutations, `r length(BHCPos)` positive CpGs with BH and `r length(BFCPos)` with Bonferroni.  

### Inverse CpGs

```{r}
permsNeg <- permsSum[permsSum$Direction == "Negative", ]$CpG
BHCNeg <- BHSum[BHSum$Direction == "Negative", ]$CpG
BFCNeg <- BFSum[BFSum$Direction == "Negative", ]$CpG

upset(fromList(list(Permutation = permsNeg, BH = BHCNeg, Bonferroni = BFCNeg)),       order.by = "freq")
```

We found `r length(permsNeg)` negative CpGs with the permutations, `r length(BHCNeg)` negative CpGs with BH and `r length(BFCNeg)` with Bonferroni.  

## TCs found

```{r}
permsTC <- unique(modU[modU$sigPair, ]$TC)
BHTC <- unique(modU[modU$sigPair_BH, ]$TC)
BFTC <- unique(modU[modU$sigPair_BF, ]$TC)

upset(fromList(list(Permutation = permsTC, BH = BHTC, Bonferroni = BFTC)), 
      order.by = "freq")
```

We found `r length(permsTC)` TCs with the permutations, `r length(BHTC)` TCs with BH and `r length(BFTC)` with Bonferroni. 

```{r}
modU <- modU %>%
  left_join(select(expAnnot, TC, Coding), by = "TC")

permsCodTC <- unique(modU[modU$sigPair & modU$Coding == "coding", ]$TC)
BHCodTC <- unique(modU[modU$sigPair_BH & modU$Coding == "coding", ]$TC)
BFCodTC <- unique(modU[modU$sigPair_BF & modU$Coding == "coding", ]$TC)

upset(fromList(list(Permutation = permsCodTC, BH = BHCodTC, Bonferroni = BFCodTC)), order.by = "freq")
```

We found `r length(permsCodTC)` coding TCs with the permutations, `r length(BHCodTC)` coding TCs with BH and `r length(BFCodTC)` with Bonferroni. 

